---
title: "README"
author: "yutian"
date: "3/28/2023"
output: github_document
header-includes: 
- \usepackage{graphicx}
- \usepackage{float}
- \usepackage{subfig}
---

# Introduction

The Immune Variation (ImmVar) project assayed gene expression in CD14 +CD16 − monocytes and CD4 + T-cells on the Affymetrix Human Gene 1.0 ST Array platform in order to characterize the role of cell type in genetic regulation of gene expression [1]. 

# Preparation

## Install/Load libraries & Load data 

```{r,message=FALSE}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

if (!require("variancePartition", quietly = TRUE))BiocManager::install("variancePartition")
if (!require("variancePartition", quietly = TRUE))BiocManager::install("dplyr")

library(variancePartition)
library(dplyr)
```

## Prepare Inputs for VariancePartition

As the inputs for VariancePartition, there are three components to an analysis:

- 1. Gene expression data
- 2. Meta- data about each experiment
- 3. Formula indicating which meat-data variables to consider

### 1. Gene expression data

```{r}
exprObj<- readRDS("data/exprObj.rds")
head(exprObj)
```


`exprObj` is a expression set, there are 33297 features and 984 samples

```{r}
#Extract expression data from `exprObj`
geneExpr<- exprObj@assayData$exprs
```

### 2. Meta Data

It should be a dataframe with information about each experiment.
Here, we have the `info` dataframe, with information about 42 variables including age, batch, sex and so on. We subset this dataframe to include variables we are interested in only. 

```{r}
info <- readRDS("data/info.rds")
#Subset dataframe & save it as `info_in` 
info_in <- info %>% select(cellType, Individual, Age, Sex, Batch)
head(info_in)
```

### 3. Formula indicating which meat-data variables to consider

The analysis is built on top of the `lme4` package

Variables must not be a continuous variable to be a fixed effect.

In this study,`Age`  is a continuous variable and it is a fixed effect and `Batch, Sex, Individual, cellType` are random effects. In the random effect, the "|" is set before and after the random slope and random intercept.


```{r}
form <- ~ Age + (1|Sex) +  (1|Individual) + (1|Batch) + (1|cellType)
```

#### Check data (Dimension & Missing Values)

```{r}
dim(info)[1] == dim(geneExpr)[2] # number of objects 
sum(is.na(info)) == 0
sum(is.na(geneExpr)) ==0
```

# Fit model & extract results Before Batch effect removal

```{r}
# Fit model and extract results
# 1) fit linear mixed model on gene expression
# If categorical variables are specified,
# a linear mixed model is used
# If all variables are modeled as fixed effects,
# a linear model is used
# each entry in results is a regression model fit on a single gene
# 2) extract variance fractions from each model fit
# for each gene, returns fraction of variation attributable
# to each variable
# Interpretation: the variance explained by each variables
# after correcting for all other variables
# Note that geneExpr can either be a matrix,
# and EList output by voom() in the limma package,
# or an ExpressionSet
```

Since the `fitExtractVarPartModel` takes a long time to run, I have saved it to 
`varPart.csv` after running it at the first time , and we could read it every time afterwards

### run it at the first time 

```{r}
if (!file.exists("data/varPart.csv"))
varPart <- fitExtractVarPartModel( geneExpr, form, info )
#write.csv(data.frame(varPart), "../data/varPart.csv")
```

### run it every time afterwards

```{r}
if (file.exists("data/varPart.csv"))
varPart <- read.csv("data/varPart.csv",row.names = 1)
```

```{r}
# sort variables (i.e. columns) by median fraction
# of variance explained
vp <- sortCols( varPart )
par(mfrow = c(1, 2))
# Figure 1a
# Bar plot of variance fractions for the first 10 genes
plotPercentBars( vp[1:10,] )
# Figure 1b
# violin plot of contribution of each variable to total variance
plotVarPart( vp )
```

**Figure 1a : This bar plot plots partition results for 10 subset genes before batch effect removed**

Based on Figure 1a, it is evident that the variable Batches accounts for more than 30% of the variation in expression before correcting for batches. Gene expression studies often have substantial batch effects that can confound biological variables. Therefore, batch correction is necessary to remove batch effects with a linear mixed model .

**Figure 1b : This violin plot plots partition results for 10 subset genes before batch effect removed**

Figure 1b indicates the same thing as Figure 1a, which is Batches explained the most of the variance. Batch effects should be removed before fitting model.

# Removing batch effects before fitting model

As mentioned, the `fitExtractVarPartModel` takes a long time to run, I have saved the one after removed batch effects as "varPartResid_Matrix.Rds" after running it at the first time. This way, you can read it every time afterwards without having to run the function again.

### run it at the first time 

```{r}
if (!file.exists("data/varPartResid_Matrix.Rds"))
{# extract residuals directly without storing intermediate results
residList <- fitVarPartModel( geneExpr, ~ (1|Batch), info,
fxn=residuals )
# convert list to matrix
residMatrix = do.call(rbind, residList)
varPartResid_Matrix <- fitExtractVarPartModel( residMatrix, form, info_in )
saveRDS(varPartResid_Matrix,"data/varPartResid_Matrix.Rds")}
```

### run it every time afterwards

```{r}
if (file.exists("data/varPartResid_Matrix.Rds"))
varPartResid_Matrix <- readRDS("~/Desktop/Interviewtest/variancePartition/data/varPartResid_Matrix.Rds")
```

# Extract results After Batch effect removal

```{r,warning=FALSE}

vp_Resid_Matrix <- sortCols( varPartResid_Matrix )
par(mfrow = c(1, 2))
# Figure 2a
# Bar plot of variance fractions for the first 10 genes
plotPercentBars( vp_Resid_Matrix [1:10,] )
# Figure 2b
# violin plot of contribution of each variable to total variance
plotVarPart( vp_Resid_Matrix  )
```

**Figure 2a,2b: Barplot and Violin plot of the distribution of variance explained by each variable across all genes after batch effect removed**

After removing batch effects, it was observed that `Cell Type` explains most of the variances followed by Individual. The effects of `Age`, `Sex`, and `Batch` on explaining the variance are negligible. `Residuals` refers to anything that cannot be explained by the model.

## Plot expression stratified by other variables 

`plotStratify` plots the expression of a gene stratified by
the specified variable. Here, I am going to a gene expression
trait stratified by `cellType` (Figure 3a) or `Individual` (Figure 3b)



```{r}
# get gene with the highest variation across Celltype
# create data.frame with expression of gene i and celltype
# type for each sample
i <- which.max( varPartResid_Matrix$cellType )
GE <- data.frame( Expression = geneExpr[i,], cellType= info_in$cellType)
# Figure 3a
# plot expression stratified by celltype
plotStratify( Expression ~ cellType, GE, main=rownames(geneExpr)[i])
#
# get gene with the highest variation across Individuals
# create data.frame with expression of gene i and celltype
# type for each sample
i <- which.max( varPartResid_Matrix$Individual )
GE <- data.frame( Expression = geneExpr[i,],
Individual = info_in$Individual)
# Figure 3b
# plot expression stratified by individuals
label <- paste("Individual:", format(varPart$Individual[i]*100,
digits=3), "%")
main <- rownames(geneExpr)[i]
plotStratify( Expression ~ Individual, GE, colorBy=NULL,
text=label, main=main)
```

**Figure 3a: Plot gene expression stratified by a) cellType**

Gene 8130499, also known as gene DYNLT1 based on the probe annotation, shows the most variance in Cell Type and is more expressed in Monocytes compared to CD4 T-cells.

**Figure 3b: Plot gene expression stratified by b) Individual**

Gene 7903765, also known as gene GSTM1, shows the most variance among all individuals. 

```{r}
mean(varPartResid_Matrix$Individual)
mean(varPartResid_Matrix$cellType)
```

## Assess correlation between all pairs of variables

```{r, warning=FALSE}
# Compute Canonical Correlation Analysis (CCA)
# between all pairs of variables
# returns absolute correlation value
form_1 <- ~ Age + (1|Batch) + (1|cellType)+ (1|Individual)
C = canCorPairs( form_1, info_in)
# Plot correlation matrix
plotCorrMatrix( C )

```

**Figure 4: Assess correlation between all pairs of variables**

It can be seen that `Individual` is highly correlated with `Age`, `Batch` and slightly lower correlated with `cellType`


# Results Interpretation

```{r}
mean(vp_Resid_Matrix$cellType)
mean(vp_Resid_Matrix$Individual)
```

Here, we have analyzed 984 objects from both cell types and it reveals that multiple variables contribute to expression variation in this dataset (Fig. 1 a). After removing batch effect,  it is apparent that the variation across cell types is the strongest biological driver of variation (22.1%) followed by variation across individuals (8.2%)(Figure 2). Variation across batches is negligible after correcting for variation due to other variables. Besides, the effects of age and sex are also negligible after correcting for other variables.




